The Impact of Accretion Disk Winds on the Optical Spectra of Cataclysmic Variables

### 2015-04-08 draft.tex
bringing up to final submission

\begin{figure}	%fullpage
\centering
\includegraphics[width=0.7\textwidth]{figures/fig1.eps}
\caption{
Optical spectra of three nova-like variables: 
RW Sex (top; Beuermann et al. 1992),
IX Vel (top middle; A. F. Pala \& B. T. Gaensicke, private communication) 
and RW Tri in and out of eclipse (bottom two panels; Groot et al. 2004).
The data for RW Sex and RW Tri were digitized from the respective publications,
and the IX Vel spectrum was obtained using the XSHOOTER spectrograph 
on the Very Large Telescope on 2014 October 10.
These systems have approximate inclinations of $30^\circ$, $60^\circ$ and $75^\circ$ 
(see section 5.4) respectively. 
The trend of increasing Balmer line emission with inclination can be seen.
In RW Tri strong single-peaked emission in the Balmer lines is seen even
in eclipse, indicating that the lines may be formed in a spatially
extensive disk wind, and there is even a suggestion 
of a (potentially wind-formed) recombination continuum in the eclipsed
spectrum. We have attempted to show each spectrum over a similar dynamic range.
}
\label{novalikes}
\end{figure}

@@

In our implementation of the macro-atom approach, we also explicitly
take into account the photoionization and collisional ionization rates
between a lower level, $l$, and the continuum (or, in the case of ions
with more than one bound electron, the ground state of the upper ion),
$\kappa$,
\begin{equation}
{\cal R}_{l \kappa}= n_l \int_{\nu_0}^{\infty} \frac{ 4 \pi J_{\nu}
  \sigma_{\nu}}{h \nu} d\nu + C_{l \kappa} n_l n_e.
\end{equation}
Here, $\sigma_{\nu}$ is the photoionization cross section, and $J_{\nu}$
is the mean intensity. The corresponding recombination rate is given
by 
\begin{equation}
{\cal R}_{\kappa l} = \alpha_{\kappa l} n_{\kappa} n_e + C_{\kappa l}
n_\kappa n_e, \\
\end{equation}
where $\alpha_{\kappa l}$ is the radiative recombination coefficient
to level $l$. This treatment means that radiative and collisional
rates to and from all levels are considered when calculating both the
ionization state and the level populations, although we neglect 
ionization directly to excited levels of the upper ion. The
\cite{vanregemorter} approximation is used for collisional
transitions. This means that collisions between radiatively
forbidden transitions are not taken into account when one 
splits levels ito $l$- and $s$-subshells, as well
as principal quantum number, $n$ (as we have done with He~\textsc{i}; 
see section~\ref{sec:data}). Although this approximation is, in general, 
a poor one, the effect is second order in the physical 
regime where recombination lines are formed in our models. 
This is because bound-free processes are dominant in determining 
level populations and emissivities. We have verified that this 
is indeed the case in the He~\textsc{i} emission regions in our models.

@@

SDL05 implemented a macro-atom treatment of H in \py\ and used
this to predict the observable properties of a pure H wind
model for YSOs. Our goal here is to simultaneously model the optical
and ultraviolet spectra of high-state CVs. Since the optical spectra
are dominated by H and He recombination lines, both of these species
need to be treated as macro-atoms. The UV spectra, on the other hand,
are dominated by resonance lines associated with metals. This means we
need to include these species in our models, but they can be treated 
with our (much faster) simple-atom approach. We have therefore
implemented a hybrid ionization and excitation scheme into \py. Any
atomic species can now be treated either in our simple-atom
approximation or with the full macro-atom machinery. In our CV
models, we treat H and He as macro-atoms and all metals as
simple-atoms. Species treated with either method
are full taken into account as sources of both bound-free opacity and line opacity, 
and contribute to the heating and cooling balance of the plasma.

@@

Our main goal is to test whether the type of disk wind model that has
been successful in explaining the UV spectra of CVs could also have a
significant impact on the optical continuum and emission line spectra
of these systems. In order to set a benchmark, we therefore begin by
investigating one of the fiducial CV wind models that was used by SV93
and LK02 to simulate the UV spectrum of a typical high-state
system. The specific parameters for this model (model A) are listed in
Table~1. A key point is that the wind mass-loss rate in this model is
set to 10$\%$ of the accretion rate through the disk. The inner edge of
the wind ($r_{min}$) is set to $4~R_{WD}$ following SV93. 
The sensitivity to some of these parameters is briefly discussed in
section~5.

@@

\subsection{Sensitivity to Parameters}

This revised model demonstrates that one can achieve a more
realistic optical spectrum by altering just two kinematic parameters. 
However, it may also be possible to achieve this by modifying
other free parameters such as $\dot{M}_{wind}$, the opening angles of the wind and the 
inner and outer launch radii. For example, increasing the mass loss rate of the wind
increases the amount of recombination emission (which scales as $\rho^2$), 
as well as lowering the ionization parameter and increasing optical depth through the wind. 
Wider launch radii and opening angles lead to a larger emitting volume, but this is moderated by a decrease in density 
for a fixed mass-loss rate. We also note that the inner radius of $4~R_{WD}$ adopted by SV93 
affects the emergent UV spectrum seen at inclinations $<\theta_{min}$ as 
the inner disk is uncovered. This causes less absorption in the UV resonance lines,
but the effect on the optical spectrum is negligible.
We have verified this general behaviour, but
we suggest that future work should investigate the effect of these parameters in more detail
as well as incorporating a treatment of clumping.
If a wind really does produce the line and continuum emission seen in optical spectra of high-state CVs, then
understanding the true mass loss rate and geometry of the outflow is clearly important.

>>>>

\begin{figure}	%fullpage
\centering
\includegraphics[width=0.7\textwidth]{figures/fig1.eps}
\caption{
Optical spectra of three nova-like variables: 
RW Sex (top; Beuermann et al. 1992),
IX Vel (top middle; A. F. Pala \& B. T. Gaensicke, private communication) 
and RW Tri in and out of eclipse (bottom two panels; Groot et al. 2004).
The data for RW Sex and RW Tri were digitized from the respective publications,
and the IX Vel spectrum was obtained using the XSHOOTER spectrograph 
on the Very Large Telescope on 2014 October 10.
These systems have approximate inclinations of $30^\circ$, $60^\circ$ and $80^\circ$ 
(see section 5.4) respectively. 
The trend of increasing Balmer line emission with inclination can be seen.
In RW Tri strong single-peaked emission in the Balmer lines is seen even
in eclipse, indicating that the lines may be formed in a spatially
extensive disk wind, and there is even a suggestion 
of a (potentially wind-formed) recombination continuum in the eclipsed
spectrum. We have attempted to show each spectrum over a similar dynamic range.
}
\label{novalikes}
\end{figure}

@@

In our implementation of the macro-atom approach, we also explicitly
take into account the photoionization and collisional ionization rates
between a lower level, $l$, and the continuum (or, in the case of ions
with more than one bound electron, the ground state of the upper ion),
$\kappa$,
\begin{equation}
{\cal R}_{l \kappa}= n_l \int_{\nu_0}^{\infty} \frac{ 4 \pi J_{\nu}
  \sigma_{\nu}}{h \nu} d\nu + C_{l \kappa} n_l n_e.
\end{equation}
Here, $\sigma_{\nu}$ is the photoionization cross section, and $J_{\nu}$
is the mean intensity. The corresponding recombination rate is given
by 
\begin{equation}
{\cal R}_{\kappa l} = \alpha_{\kappa l} n_{\kappa} n_e + C_{\kappa l}
n_\kappa n_e, \\
\end{equation}
where $\alpha_{\kappa l}$ is the radiative recombination coefficient
to level $l$. This treatment means that radiative and collisional
rates to and from all levels are considered when calculating both the
ionization state and the level populations, although we neglect 
ionization directly to excited levels of the upper ion. The
\cite{vanregemorter} approximation is used for collisional
transitions. This means that collisions between radiatively
forbidden transitions are not taken into account when one 
splits levels into $l$- and $s$-subshells, as well
as principal quantum number, $n$ (as we have done with He~\textsc{i}; 
see section~\ref{sec:data}). Although this approximation is, in general, 
a poor one, the effect is second order in the physical 
regime where recombination lines are formed in our models. 
This is because bound-free processes are dominant in determining 
level populations and emissivities. We have verified that this 
is indeed the case in the He~\textsc{i} emission regions in our models.

@@

SDL05 implemented a macro-atom treatment of H in \py\ and used
this to predict the observable properties of a pure H wind
model for YSOs. Our goal here is to simultaneously model the optical
and ultraviolet spectra of high-state CVs. Since the optical spectra
are dominated by H and He recombination lines, both of these species
need to be treated as macro-atoms. The UV spectra, on the other hand,
are dominated by resonance lines associated with metals. This means we
need to include these species in our models, but they can be treated 
with our (much faster) simple-atom approach. We have therefore
implemented a hybrid ionization and excitation scheme into \py. Any
atomic species can now be treated either in our simple-atom
approximation or with the full macro-atom machinery. In our CV
models, we treat H and He as macro-atoms and all metals as
simple-atoms. Species treated with either method
are fully taken into account as sources of both bound-free opacity and line opacity, 
and contribute to the heating and cooling balance of the plasma.

@@

Our main goal is to test whether the type of disk wind model that has
been successful in explaining the UV spectra of CVs could also have a
significant impact on the optical continuum and emission line spectra
of these systems. In order to set a benchmark, we therefore begin by
investigating one of the fiducial CV wind models that was used by SV93
and LK02 to simulate the UV spectrum of a typical high-state
system. The specific parameters for this model (model A) are listed in
Table~1. A key point is that the wind mass-loss rate in this model is
set to 10$\%$ of the accretion rate through the disk. We follow SV93
in setting the inner edge of the wind ($r_{min}$) to $4~R_{WD}$. 
The sensitivity to some of these parameters is briefly discussed in
section~5.

@@

\subsection{Sensitivity to Model Parameters}

This revised model demonstrates that one can achieve a more
realistic optical spectrum by altering just two kinematic parameters. 
However, it may also be possible to achieve this by modifying
other free parameters such as $\dot{M}_{wind}$, the opening angles of the wind and the 
inner and outer launch radii. For example, increasing the mass loss rate of the wind
increases the amount of recombination emission (which scales as $\rho^2$), 
as well as lowering the ionization parameter and increasing optical depth through the wind. 
Larger launching radii and covering factors tend to lead to a larger emitting volume, 
but this is moderated by a decrease in density 
for a fixed mass-loss rate. We also note that the inner radius of $4~R_{WD}$ adopted by SV93 
affects the emergent UV spectrum seen at inclinations $<\theta_{min}$ as 
the inner disk is uncovered. This causes less absorption in the UV resonance lines,
but the effect on the optical spectrum is negligible.
We have verified this general behaviour, but
we suggest that future work should investigate the effect of these parameters in more detail,
as well as incorporating a treatment of clumping.
If a wind really does produce the line and continuum emission seen in optical spectra of high-state CVs, then
understanding the true mass loss rate and geometry of the outflow is clearly important.
